---
title: "Further applications of Paralog Annotation"
# output: rmarkdown::github_document
output: 

  # word_document:
  #  reference_docx: template.docx

  pdf_document:
    toc: true
    number_sections: true
    fig_caption: yes
    
editor_options: 
  chunk_output_type: inline
bibliography: bibliography.bib
always_allow_html: yes
---

<!--Load Packages and function-->
<!-- ```{r echo=FALSE, message=FALSE, warning=FALSE, include=FALSE, results='hide'} -->
```{r echo=FALSE, message=FALSE, warning=FALSE, results='hide',include=FALSE}
Packages = c("tidyverse", "plyr", "dplyr", "httr", "jsonlite","XML")
new.packages = Packages[!(Packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages, repos = "https://cran.ma.imperial.ac.uk/")
# lapply(Packages, library, character.only = TRUE)
library("plyr")
library("dplyr")
library("tidyverse")
library("httr")
library("jsonlite")
library("XML")
Packages = c("ggplot2", "ggsignif", "knitr", "png", "grid", "tinytex", "pander", "kableExtra", "DiagrammeR")
new.packages = Packages[!(Packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages, repos = "https://cran.ma.imperial.ac.uk/")
lapply(Packages, library, character.only = TRUE)
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
bioconductor_Packages = c("biomaRt", "org.Hs.eg.db")
new.packages = bioconductor_Packages[!(bioconductor_Packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) BiocManager::install(new.packages)
lapply(bioconductor_Packages, library, character.only = TRUE)
```
```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.pos = 'H')
```
\newpage
# To do list


# Introduction

# Material and Methods

## Datasets
Variants with CLNSIG=Pathogenic, CLNSIG=Likely_pathogenic, CLNSIG=Pathogenic/Likely_pathogenic grepped out of clinvar_20190114_GRCh37.vcf
Note there are a few variants present in GRCh38 but not in GRCh37 build. Take only missense.

```{r, cache=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
xB = read.csv("/media/nick/Data/PhD/Paralogues/ParalogueAnnotation_personal/data/clinvar/clinvar_20190114_GRCh38_onlyBenign_and_Likely_benign.vcf", sep = "\t", stringsAsFactors = F, header = F, comment.char = "#")
yB = read.csv("/media/nick/Data/PhD/Pfams/Pfam_meta_domains/data/clinvar/clinvar_20190114/clinvar_20190114_GRCh37_onlyBenign_and_Likely_benign.vcf", sep = "\t", stringsAsFactors = F, header = F, comment.char = "#")
xB1 = xB[grepl("missense_variant",xB$V8),]
yB1 = yB[grepl("missense_variant",yB$V8),]
dif_id = setdiff(xB1$V3, yB1$V3)

xP = read.csv("/media/nick/Data/PhD/Paralogues/ParalogueAnnotation_personal/data/clinvar/clinvar_20190114_GRCh38_onlyPathogenic_and_Likely_pathogenic.vcf", sep = "\t", stringsAsFactors = F, header = F, comment.char = "#")
yP = read.csv("/media/nick/Data/PhD/Pfams/Pfam_meta_domains/data/clinvar/clinvar_20190114/clinvar_20190114_GRCh37_onlyPathogenic_and_Likely_pathogenic.vcf", sep = "\t", stringsAsFactors = F, header = F, comment.char = "#")
xP1 = xP[grepl("missense_variant",xP$V8),]
yP1 = yP[grepl("missense_variant",yP$V8),]
dif_id = setdiff(xP1$V3, yP1$V3)
```

For all missense synthetic exome run on HPC, used qsub_scripts:
```{r eval = FALSE}
#after each run CHECK with bash command in USEFUL CODE TIPS in Paralog_annotation.Rmd

#Input files located in /rds/general/project/lms-ware-analysis/live/nick/RBH-work/Paralog_Anno/data_files/all_possible_mutation/synthetic_exome/synthetic_exome_GRCh37_renamed/
#Plan is to run tableize for pfam analysis on the .out_paraloc files and go from there.

HPC_run_chrom1 = data.frame(
  file = c(
"0-50",
"50-99",
"100-360"
),
done = c(1,0,0)
)

HPC_run_chrom2 = data.frame(
  file = c(
"0-9",
"10-99", #93 had warnings
"100-263"
),
done = c(0,0,0)
)

HPC_run_chrom3 = data.frame(
  file = c(
"0-9",
"10-99",
"100-204"
),
done = c(0,0,0)
)

HPC_run_chrom4 = data.frame(
  file = c(
"0-9",
"10-99",
"100-139"
),
done = c(0,0,0)
)

HPC_run_chrom5 = data.frame(
  file = c(
"0-9",
"10-99", 
"100-159"
),
done = c(0,0,0)
)

HPC_run_chrom6 = data.frame(
  file = c(
"0-9",
"10-99", 
"100-179"
),
done = c(0,0,0)
)

HPC_run_chrom7 = data.frame(
  file = c(
"0-9",
"10-99",
"100-172"
),
done = c(0,0,0)
)

HPC_run_chrom8 = data.frame(
  file = c(
"0-9",
"10-99",
"100-123"
),
done = c(0,0,0)
)

HPC_run_chrom9 = data.frame(
  file = c(
"0-9",
"10-99",
"100-144"
),
done = c(0,0,0)
)

HPC_run_chromY = data.frame(
  file = c(
"0-11"
),
done = c(1)
)
```



## Annotation of variants and transfer of annotations across homologous pfam positions

The easiest thing to do first is to run VEP to annotate variant with protein and codon info and then run tableize for easy info extraction

e.g. Run VEP:
```{bash eval = FALSE, include = TRUE}
#PBS -lwalltime=72:0:0
#PBS -lselect=1:ncpus=1:mem=16gb

module load anaconda3/personal
source /home/nyl112/.bashrc

/home/nyl112/perl5/perlbrew/bin/perlbrew switch perl-5.22.0

perl /work/nyl112/ensembl-vep/vep --force_overwrite --vcf --allele_number --canonical --offline --cache --dir_cache /work/nyl112/ -assembly GRCh37 --port 3337 -i /work/nyl112/Pfams/Pfam_meta_domains/data/clinvar/clinvar_20190114/clinvar_20190114_GRCh37_onlyPathogenic_and_Likely_pathogenic.vcf -o /work/nyl112/Pfams/Pfam_meta_domains/data/clinvar/clinvar_20190114/clinvar_20190114_GRCh37_onlyPathogenic_and_Likely_pathogenic.out_no_plugin
```


e.g. Run tableize:
```{bash eval = FALSE, include = TRUE}
#PBS -lwalltime=72:0:0
#PBS -lselect=1:ncpus=1:mem=5gb

module load anaconda3/personal
source /home/nyl112/.bashrc

python2 /work/nyl112/loftee/src/tableize_vcf.py --vcf /work/nyl112/Pfams/Pfam_meta_domains/data/clinvar/clinvar_20190114/clinvar_20190114_GRCh37_onlyPathogenic_and_Likely_pathogenic.out_no_plugin --out /work/nyl112/Pfams/Pfam_meta_domains/data/clinvar/clinvar_20190114/clinvar_20190114_GRCh37_onlyPathogenic_and_Likely_pathogenic.out_no_plugin_tableized_org --do_not_minrep --include_id --vep_info SYMBOL,Protein_position,Amino_acids,Codons,BIOTYPE --split_by_transcript --canonical_only

python2 /work/nyl112/loftee/src/tableize_vcf.py --vcf /work/nyl112/Pfams/Pfam_meta_domains/data/clinvar/clinvar_20190114/clinvar_20190114_GRCh37_onlyBenign_and_Likely_benign.out_no_plugin --out /work/nyl112/Pfams/Pfam_meta_domains/data/clinvar/clinvar_20190114/clinvar_20190114_GRCh37_onlyBenign_and_Likely_benign.out_no_plugin_tableized_org --do_not_minrep --include_id --vep_info SYMBOL,Protein_position,Amino_acids,Codons,BIOTYPE --split_by_transcript --canonical_only
```


## Statistical measures
For a pathogenic paralogue alignment:

true positive (TP) = pathogenic query variant with a paralogous pathogenic hit;

false positive (FP) = benign query variant with a paralogous pathogenic hit;

false negative (FN) = pathogenic query variant with no paralogous pathogenic hit;

and true negative (TN) = benign query variant with no paralogous pathogenic hit.\newline


Likewise for a benign paralogous alignment:

TP = benign query variant with a paralogous benign hit; 

FP = pathogenic query variant with a paralogous benign hit; 

FN = benign query variant with no paralogous benign hit; 

and TN = pathogenic query variant with no paralogous benign hit.\newline

Positive Predictive Values (PPV) and Sensitivties are therefore calculated by:

$$PPV = \frac{TP}{TP+FP}$$
$$Sensitivity = \frac{TP}{TP+FN}$$

P values are calculated via a Fisher's exact test on a 2x2 contingency table tabulating the number of pathogenic and benign variants of interest and how many of those are predicted as pathogenic or benign.

## Calculation of Etiological Fractions

Odds ratios (OR) are calculated by:

$$ OR =  \frac{(a/b)}{(c/d)} $$

where:

$a =$ number of variants predicted to be pathogenic by paralogue annotation in the diseased cohort

$b =$ number of varaints predicted to be pathogenic by paralogue annotation in the control cohort

$c =$ number of variants not predicted to be pathogenic by paralogue annotation in the diseased cohort

$d =$ number of variants not predicted to be pathogenic by paralogue annotation in the control cohort

and the 95% confidence intervals for OR values are calculated according to @altman1991practical via:

$$ 95\% \ CI = [e^{\ln(OR) - 1.96 \cdot SE(\ln(OR))},e^{\ln(OR) + 1.96 \cdot SE(\ln(OR))}] $$
where:

$$ SE(\ln(OR)) =  \sqrt{\frac{1}{a}+\frac{1}{b}+\frac{1}{c}+\frac{1}{d}} $$

Etiological Fractions (EF) can then be calculated by:

$$ EF = \frac{OR - 1}{OR} $$
and the 95% confidence intervals for EF values are calculated according to @hildebrandt2006calculating via:

$$ 95\% \ CI = [\widehat{EF} - z_{1-\frac{a}{2}} \sqrt{\widehat{VAR}(\widehat{\Phi})}, \ min( \widehat{EF} + z_{1-\frac{a}{2}} \sqrt{\widehat{VAR}(\widehat{\Phi})}, \ 1)]  $$
where the asymptomatic variance estimator of $\widehat{\Phi}$ is given by:

$$ \widehat{VAR}(\widehat{\Phi}) = \widehat{\Phi}^2 [\frac{1-\widehat{\pi}_{0}}{N_0\widehat{\pi}_{0}}+\frac{1-\widehat{\pi}_{1}}{N_1\widehat{\pi}_{1}}]$$
and:
$\widehat{\pi}_{1} = a/(b + a)$
$\widehat{\pi}_{0} = c/(d + c)$ 
$N_0 = c + d$ 
$N_1 = a + b$ 
$\widehat{\Phi}^2 = (\widehat{\pi}_{0}/\widehat{\pi}_{1})^2$

NOTE ACCORDING TO https://www.medcalc.org/calc/odds_ratio.php - "Where zeros cause problems with computation of the odds ratio or its standard error, 0.5 is added to all cells (a, b, c, d) (Pagano & Gauvreau, 2000; Deeks & Higgins, 2010)."

## Paralogue stats
The additional statistics were calculated by programmatically extracting the genes of interest (using `src/check_what_clinvar_genes.py` and `src/Find_unique_genes.py`) and then retrieving relevant information manually from [Ensembl's Bioimart](https://www.ensembl.org/biomart)

Alternatively, this can be reproduced using biomaRt package
                 
## Comparison to SIFT/REVEL scores
SIFT and Mcap scores were retreived using ANNOVAR with dbnsfp30a [@wang2010annovar,@liu2016dbnsfp]. Revel scores were downloaded from https://sites.google.com/site/revelgenomics/downloads [@ioannidis2016revel].

Revel did not provide scores for all available variants in the dataset and therefore to ensure the same variants were being annotated by all tools

# Results and Discussion
## PDB stats
```{r cache=TRUE, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
stats = read.csv("Data_Export_Growth.csv", stringsAsFactors = F, header = T)
stats = stats[order(stats$Year),]

cols = c( "Released structures" = "red", "Total entries" = "blue")

plot = ggplot(stats, aes(x=Year)) + 
  geom_line(aes(y=Total.Number.of.Entries.Available, color = "Total entries")) +
  geom_line(aes(y=Number.of.Structures.Released.Annually, color = "Released structures")) + 
  scale_color_manual(values = cols, guide = guide_legend(reverse = TRUE)) + 
  theme(aspect.ratio = 2/1,
        legend.title = element_blank(),
        # legend.key.size = unit(5,"line"),
        # legend.position = c(0.77,0.89),
        legend.text = element_text(size=14),
        axis.text = element_text(size=14),
        axis.text.x = element_text(colour = "black"),
        axis.title.y = element_text(size=20),
        axis.title.x = element_text(size=20)) +
  # labs(color = "Legend") +
  # theme(axis.text.x=element_blank(), axis.ticks = element_blank()) + 
  ylab("Number of Entries") +xlab("Year") 
  # theme(legend.title = element_blank())
ggsave("images/PDB_stats.png")
```

## Counting unique variant locations in protein for pymol
```{r cache=TRUE, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
clinvar_P_LP_tableized_for_pfam_analysis = read.csv(file="/media/nick/821ED5711ED55F2B/Users/Nick/Documents/PhD/Pfams/Pfam_meta_domains/data/clinvar/clinvar_20190114/clinvar_20190114_GRCh37_onlyPathogenic_and_Likely_pathogenic.out_no_plugin_tableized_for_pfam_analysis", sep = "\t", header=TRUE, stringsAsFactors=FALSE)
SCN5A_unique_variant_pos = unique(clinvar_P_LP_tableized_for_pfam_analysis[clinvar_P_LP_tableized_for_pfam_analysis$SYMBOL == "SCN5A","Protein_position"])
length(SCN5A_unique_variant_pos)
SCN2A_unique_variant_pos = unique(clinvar_P_LP_tableized_for_pfam_analysis[clinvar_P_LP_tableized_for_pfam_analysis$SYMBOL == "SCN2A","Protein_position"])
length(SCN2A_unique_variant_pos)

```

## Analyzing distances between closest points
```{r cache=TRUE, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
P.distances = read.csv("Structures/Phyre_SCN5A_aligned_with_Phyre_SCN2A.prb.P_LP_distance_calculated_out", sep = ",", header = FALSE, stringsAsFactors = FALSE)
P.distances$V4 = paste(P.distances$V1, P.distances$V2, sep = " ")
P.distances$V5 = "Pathogenic to Pathogenic"
colnames(P.distances) = c("Query_variant", "Ref_variant", "Distance", "Query_Ref", "Diseased")
P.distances$Query_protein_position = sapply(str_split(P.distances$Query_variant, " "), tail, n=1L) 
mean(P.distances$Distance)
sd(P.distances$Distance)

B.distances = read.csv("Structures/Phyre_SCN5A_aligned_with_Phyre_SCN2A.prb.B_LB_distance_calculated_out", sep = ",", header = FALSE, stringsAsFactors = FALSE)
B.distances$V4 = paste(B.distances$V1, B.distances$V2, sep = " ")
B.distances$V5 = "Benign to Pathogenic"
colnames(B.distances) = c("Query_variant", "Ref_variant", "Distance", "Query_Ref", "Diseased")
B.distances$Query_protein_position = sapply(str_split(B.distances$Query_variant, " "), tail, n=1L) 
mean(B.distances$Distance)
sd(B.distances$Distance)

All.distances = rbind(P.distances, B.distances)

#### Pfams ####
##P/LP##
p.Unique_variant_gene_annotations = NULL
p.Unique_input_variants = NULL
files = list.files(path="/media/nick/821ED5711ED55F2B/Users/Nick/Documents/PhD/Pfams/Pfam_meta_domains/data/clinvar/clinvar_20190114/RData_objects/P_LP", pattern="*.RData", full.names=TRUE, recursive=FALSE)
for (i in files){
  print(i)
  load(i)
  if (is.null(p.Unique_variant_gene_annotations)){
    p.Unique_variant_gene_annotations = Total_annotations$Unique_variant_homologous_annotations
  } else {
  p.Unique_variant_gene_annotations = base::rbind(p.Unique_variant_gene_annotations, dplyr::setdiff(Total_annotations$Unique_variant_homologous_annotations, p.Unique_variant_gene_annotations))
  }
  if (is.null(p.Unique_input_variants)){
    p.Unique_input_variants = Total_annotations$input_variants
  } else {
    p.Unique_input_variants = base::rbind(p.Unique_input_variants, dplyr::setdiff(Total_annotations$input_variants, p.Unique_input_variants))
  }
}

SCN5A_p.Unique_variant_gene_annotations = sort(unique(p.Unique_variant_gene_annotations[p.Unique_variant_gene_annotations$Gene == "SCN5A","Protein_position.x"]))
Not_annotated = P.distances[!(P.distances$Query_protein_position %in% SCN5A_p.Unique_variant_gene_annotations),]
Not_annotated$Diseased = "Pathogenic to Pathogenic\n(removed previously annotated)"
All.distances = rbind(All.distances, Not_annotated)
All.distances$Diseased = factor(All.distances$Diseased, levels = c("Benign to Pathogenic", "Pathogenic to Pathogenic", "Pathogenic to Pathogenic\n(removed previously annotated)"))

plot = ggplot(All.distances, aes(x=Diseased, y=Distance, color=Diseased)) + 
  # geom_violin(trim = FALSE, scale = "count") + 
  # geom_boxplot() +
  # scale_fill_manual(values = c("P/LP"="#F8CECC", "B/LB"="#D5E8D4")) +
  scale_color_manual(values = c("Pathogenic to Pathogenic"="#B85450", "Benign to Pathogenic"="#82B366", "Pathogenic to Pathogenic\n(removed previously annotated)" = "#0084ff")) +
  geom_jitter(width = 0.2, alpha = 0.5) + 
  ylab("Distance (Angstroms)") +xlab("Variants") +
  theme(legend.title = element_blank(),
        legend.key.size = unit(3,"line"),
        # legend.position = c(0.77,0.89),
        legend.text = element_text(size=12),
        axis.text = element_text(size=14),
        # axis.text.x = element_text(colour = "black", angle=20, hjust = 0.5),
        axis.text.x = element_blank(),
        axis.title.y = element_text(size=20),
        axis.title.x = element_text(size=20),
        axis.ticks = element_blank()) +
  guides(color = guide_legend(override.aes = list(size = 3))) +
  stat_summary(mapping = aes(x=Diseased, y=Distance), fun.y = median, fun.ymin = median, fun.ymax = median, geom = "crossbar", width = 0.5, color = "black", fatten = 0.5) + 
  stat_summary(mapping = aes(x=Diseased, y=Distance), fun.y = function(z){quantile(z,0.25)}, fun.ymin = function(z){quantile(z,0.25)}, fun.ymax = function(z){quantile(z,0.25)}, geom = "crossbar", width = 0.2, color = "black", fatten = 0.5) +
  stat_summary(mapping = aes(x=Diseased, y=Distance), fun.y = function(z){quantile(z,0.75)}, fun.ymin = function(z){quantile(z,0.75)}, fun.ymax = function(z){quantile(z,0.75)}, geom = "crossbar", width = 0.2, color = "black", fatten = 0.5)
ggsave("images/Closest_variant_annotation.png", width = 8)
```


# References










